---
title: "Prescribing in Scotland"
output: html_document
---

```{r, message = F, echo = F, warning = F}
library(tidyverse)
library(gt)

knitr::opts_chunk$set(message = F, echo = F, warning = F)

rx_year_para <- read_csv("rx_year_para.csv")
```

```{r}
filtered_para <-
rx_year_para %>%
  filter(str_detect(bnf_chapter, "Appliance") == F) %>%
  filter(str_detect(bnf_chapter, "Dressing") == F) %>%
  filter(str_detect(bnf_chapter, "Skin") == F) %>%
  filter(str_detect(bnf_paragraph, "topical NSAIDs") == F) %>%
  filter(str_detect(bnf_paragraph, "lubricant") == F) %>%
  filter(str_detect(bnf_paragraph, "monitor") == F) %>%
  filter(str_detect(bnf_paragraph, "Local anaesthetics") == F) %>%
  filter(str_detect(bnf_paragraph, "Food") == F) %>%
  filter(total_rx_2019 >= 100000 | total_rx_2022 >= 100000) 
```

<br>

### Drugs with largest **increase** in numbers of prescriptions

```{r}

filtered_para %>%
  arrange(desc(diff)) %>%
  slice(1:10) %>%
  select(bnf_paragraph:prop_diff) %>%
  gt() %>%
  cols_label(bnf_paragraph = "Drug Class",
             total_rx_2019 = "Rx 2019",
             total_rx_2022 = "Rx 2022",
             diff = "Increase",
             prop_diff = "%") %>%
  fmt_number(c(total_rx_2019, total_rx_2022, diff),
             decimals = 0,
             use_seps = T) %>%
  fmt_percent(prop_diff,
              decimals = 0) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")

#present unfiltered too
#big total rx for medicated plasters
```

<br>

### Drugs with largest **percentage increase** in prescriptions

```{r}

filtered_para %>%
  arrange(desc(prop_diff)) %>%
  slice(1:10) %>%
  select(bnf_paragraph:prop_diff) %>%
  gt() %>%
  cols_label(bnf_paragraph = "Drug Class",
             total_rx_2019 = "Rx 2019",
             total_rx_2022 = "Rx 2022",
             diff = "Increase",
             prop_diff = "%") %>%
  fmt_number(c(total_rx_2019, total_rx_2022, diff),
             decimals = 0,
             use_seps = T) %>%
  fmt_percent(prop_diff,
              decimals = 0) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")

```

<br>

### Drugs with largest **decrease** in numbers of prescriptions

```{r}

filtered_para %>%
  arrange(diff) %>%
  slice(1:10) %>%
  select(bnf_paragraph:prop_diff) %>%
  gt() %>%
  cols_label(bnf_paragraph = "Drug Class",
             total_rx_2019 = "Rx 2019",
             total_rx_2022 = "Rx 2022",
             diff = "Decrease",
             prop_diff = "%") %>%
  fmt_number(c(total_rx_2019, total_rx_2022, diff),
             decimals = 0,
             use_seps = T) %>%
  fmt_percent(prop_diff,
              decimals = 0) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")

#present unfiltered too
#big total rx for medicated plasters
```

<br>

### Drugs with largest **percentage decrease** in prescriptions

```{r}

filtered_para %>%
  arrange(prop_diff) %>%
  slice(1:10) %>%
  select(bnf_paragraph:prop_diff) %>%
  gt() %>%
  cols_label(bnf_paragraph = "Drug Class",
             total_rx_2019 = "Rx 2019",
             total_rx_2022 = "Rx 2022",
             diff = "Decrease",
             prop_diff = "%") %>%
  fmt_number(c(total_rx_2019, total_rx_2022, diff),
             decimals = 0,
             use_seps = T) %>%
  fmt_percent(prop_diff,
              decimals = 0) %>%
  tab_options(table.align = "left",
              data_row.padding = 0,
              column_labels.font.weight = "bold")

```

<br>

### Data and Methods

This analysis includes all prescriptions in Scotland dispensed outside of hospitals. Prescriptions for appliances, food, and topical skin and eye care have been excluded.

Prescribing data were downloaded from here <https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community>

A drug class is all medication given the same paragraph code by the BNF

BNF data were downloaded from here <https://applications.nhsbsa.nhs.uk/infosystems/data/showDataSelector.do?reportId=126>

Number of prescriptions was calculated as the sum of number of items paid, as specified here <https://www.isdscotland.org/Health-topics/Prescribing-and-medicines/_docs/Open_Data_Glossary_of_Terms.pdf?1>

All code to produce this report is here <https://github.com/JessButler/scotland_prescriptions>

Any questions or comments please email [JessicaButler\@abdn.ac.uk](mailto:JessicaButler@abdn.ac.uk){.email}
